---
title: "Part 2 - Completed Version"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: github_document
always_allow_html: yes
---

## Introduction
This notebook expands mapping to a new package, `ggplot2`. We'll focus on thematic choropleth mapping with it, though you can make other types of maps as well. Along the way, we'll discussing using projected coordinate systems for mapping.

## Dependencies
This notebook requires a variety of packages for working with spatial data:

```{r load-packages}
# tidyverse packages
library(ggplot2)      # plotting

# spatial packages
library(mapview)      # preview spatial data
library(sf)           # spatial data tools

# other packages
library(here)         # file path management
library(RColorBrewer) # color palettes
```

## Working with Spatial Data Files
There are two main types of spatial data files you might run into:
  * "Shapefiles" are a type of file format (actually a collection of files) that were popularized by ESRI, the makers of the ArcGIS software platform. These files are very, very common in the GIS world. They contain both the geometric and tabular data needed to map data. They're a bit clunky to work with using your operating system - there can be over a dozen constituent files, and they all must be named identically. So, they're not the most friendly files to work with, but their ubiquity makes it important to know a bit about how to work with them.
  * "GeoJSON" files are plain text files that contain all of the core information that shapefiles contain (tabular and geometric data), though they cannot store some of the more complex metadata that shapefiles can. They're far easier to work with, however, because they are a single plain text file. 

### Reading Spatial Data Files
For data that have already been converted to geometric data, we use the `sf` package to read them. The importing process looks similar to what we used with the `.csv` file. It is the same process regardless of whether you have a shapefile or a GeoJSON file. 

We'll demonstrate this with COVID-19 data for Chicago at the ZIP code level:

```{r read-covid}
covid <- st_read(here("data", "chicago_covid.geojson"))
```

We'll still use `here()` to specify the file path, but the function is different now because we need a specialized tool for geometric data.

Now, you repeat this process for the `chicago_black_pop.geojson`, which is stored in the same directory:

```{r read-black}
black <- st_read(here("data", "chicago_black_pop.geojson"))
```

This file contains Census tract level estimates of the Black population in Chicago from the latest edition of the American Community Survey's 2015-2019 five-year estimates.

Finally, we'll read in a shapefile containing the City of Chicago's boundary:

```{r read-boundary}
city <- st_read(here("data", "chicago_boundary", "chicago_boundary.shp"))
```

Notice that we specify the `.shp` file, which is one of four files that accompany our shapefile collection.

### Previewing Data
Once we have our data read in, we can start to explore it. First, take a moment to use `View()` (or click on each of the objects in your global environment) to check out their structure and get a sense of the different columns. Next, lets use `mapview` to explore our data:

```{r preview-covid}
mapview(covid)
```

Now, you repeat this process with the Part 1 crime data for Shaw:

```{r preview-black}
mapview(black)
```

### Prepping Our Data
In order to map our data properly, we need to "transform" our data to a *projected coordinate system*. As we did in the first part, we'll use the `sf` package's `st_transform()` function to do this. Before we make the transformation, we should understand both what a projected coordinate system is and how we pick one.

Projected coordinate systems provide us with a mathematical transformation of our geographic coordinate system data, which are based on a spherical model of the earth, to a two-dimensional space. Since our computers and printed maps both render data in two-dimensions, we need to do this to deal with the distortions that come from viewing our geographic coordinate system data in two-dimensions. The key lesson here is that *we should always map in a projected coordinate system*.

There are different types of projected coordinate systems for specific parts of the world. Different systems make these mathematical transformations in different ways. So, we ultimately end up having to pick from what seems at first blush to be a relatively large menu of coordinate systems. In the United States, at the local level, most analysts select either the state plane or UTM coordinate systems. Chicago falls into the Illinois East State Plane region, whose 2007 updates are referred to using the EPSG code `3528`.

Coordinate systems are perhaps the most complex, and therefore most misunderstood, part of mapping. If you switch to mapping different regions, or change the scale (state-wide mapping, regional mapping, national mapping, global mapping), you'll have to also change the projected coordinate system you use.

To make our transformations, we'll use `st_transform()` paired with our selected EPSG code for the COVID data:

```{r transform-covid}
covid <- st_transform(covid, crs = 3528)
```
 
Now, you try this syntax out on the demographic data and:

```{r transform-black}
black <- st_transform(black, crs = 3528)
```

Finally, we need to transform our city boundary data as well:

```{r transform-city}
city <- st_transform(city, crs = 3528)
```

It is important to get all of your `sf` objects that will be mapped together into the same projected coordinate system!

## Simple Maps with `ggplot2`
### Basic Mapping of Geometric Objects
`ggplot2` is the premier graphics package for `R`. It is an incredibly powerful visualization tool that increasingly supports spatial work and mapping. The basic `ggplot2` workflow requires chaining together functions with the `+` sign. 

We'll start by creating a `ggplot2` object with the `ggplot()` function, and then adding a "geom", which provides `ggplot2` instructions on how our data should be visualized. We can read these like paragraphs:

1. First, we create an empty `ggplot2` object, **then**
2. we add the `covid` data and visualize its geometry.

```{r ggplot2-covid}
ggplot() +
  geom_sf(data = covid, fill = "#bababa")
```

The color abbreviation we use is called a "hex decimal" - it is a symbolic representation of a color model that helps us pick specific colors for our plots. I rely on the website [ColorHexa](http://colorhexa.com) to help select hex decimal values.

We can also add the `city` layer on top to give the city border a pronounced outline. `ggplot2` relies on layering different geoms to produce complicated plots. We can assign each geom a specific set of aesthetic characteristics and use data from different objects.

```{r ggplot2-covid-2}
ggplot() +
  geom_sf(data = covid, fill = "#bababa") +
  geom_sf(data = city, fill = NA, color = "#000000", size = .75)
```

These ZIP code boundaries were created by the City of Chicago, and represent generalizations of where ZIP codes are located. The challenge with working with ZIP codes is that they are not actually polygons on a map, but rather lines that represent distinct carrier routes. Because of this, USPS does not publish an authoritative shapefile of ZIP code boundaries. Some places, like Chicago, make their own. In many parts of the country, we use the Census Bureau's ZIP Code Tabulation Areas (ZCTAs) instead. There are many challenges to working with ZIP code data, and they should be used as a geography of last resort. Unfortunately, they are also the most common data we can collect from individuals because most people know the ZIP code they live in. This tension has been written about extensively in geography. 

Now it is your turn - re-create this process but map zip codes using the `black` data and use the city boundaries in `city`:

```{r ggplot2-black}
ggplot() +
  geom_sf(data = black, fill = "#bababa") +
  geom_sf(data = city, fill = NA, color = "#000000", size = .75)
```

Notice that our data don't line up well. This is one of the complexities of working with Census data. Spatial data that they provide are at the county-level, so in this case we have Cook County's Census tracts. Census tract boundaries do not always respect municipal boundaries, and so some Census tracts will cross between Chicago and neighboring municipalities.

### Mapping Quantities with `ggplot2`
If we wanted to start to map data instead of just the geometric properties, we would specify an "aesthetic mapping" using `mapping= aes()` in the geom of interest. Here, we create a fill that is the product of taking the 2019 Black population's rate. 

We always want to map per capita rates, and not the raw counts. If we do not have a population denominator, like with total population, we can map density instead by using area as our denominator. Either approach is acceptable, but it is critical that you use some form of normalization. Otherwise, you may simply find yourself making a map of where people live, instead of your phenomenon of interest. Alternatively, you may make a map that draws readers' eyes towards the largest geographic areas. Neither is a good outcome! 

We provide additional instructions about how our data should be colored with the `scale_fill_distiller()` function, which gives us access to the `RColorBrewer` palettes.

```{r ggplot2-black-choro}
## create plot
p1 <- ggplot() +
  geom_sf(data = black, mapping = aes(fill = black_rate), size = .05) +
  geom_sf(data = city, fill = NA, color = "#000000", size = .75) +
  scale_fill_distiller(palette = "Greens", trans = "reverse", name = "Population per 1,000") +
  labs(
    title = "Per Capita Black Population (2019)",
    subtitle = "Tracts in Cook County, IL",
    caption = "Map by Christopher Prener, Ph.D."
  ) +
  theme_minimal() 

## print plot
p1
```

Replicate this process, using the `case_rate` or `death_rate` column in `covid` to plot the already normalized numbers of cases per 100,000 people in each zip code:

```{r ggplot2-covid-choro}
## create plot
p2 <- ggplot() +
  geom_sf(data = covid, mapping = aes(fill = case_rate), size = .25) +
  geom_sf(data = city, fill = NA, color = "#000000", size = .75) +
  scale_fill_distiller(palette = "RdPu", trans = "reverse", name = "Cases Per 1,000 People") +
  labs(
    title = "Reported COVID-19 Cases",
    subtitle = "ZIP Codes in Chicago, IL",
    caption = "Map by Christopher Prener, Ph.D."
  ) +
  theme_minimal() 

## print plot
p2
```

## Saving Maps
To save our maps, we can use the `ggsave()` function. We'll refer to the plot object we created above, `p1`:

```{r save-race}
ggsave(filename = here("results", "chicago_black_pop.png"), plot = p1)
```

Be aware of the other arguments for `ggsave()`, which include the ability to control the dimensions and dpi of your output.

Now it's your turn - save the COVID map that you created above:

```{r save-covid}
ggsave(filename = here("results", "chicago_covid_cases.png"), plot = p2)
```
