---
title: "Part 1 - Working Version"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: github_document
always_allow_html: yes
---

## Introduction
This notebook provides a walk-through of the code for the first part of our lesson. We'll cover reading in tabular data, projecting it so that it can be mapped, and create some simple interactive maps using the `leaflet` JavaScript library along the way. We'll also talk about geographic coordinate systems.

## Dependencies
This notebook requires the following packages:

```{r load-packages}
# tidyverse packages
library(magrittr)     # pipe operator
library(readr)        # csv tools

# mapping packages
library(mapview)      # preview spatial data
library(leaflet)      # interactive maps
library(sf)           # spatial tools

# other packages
library(here)         # file path management
```

## Leaflet
As a way to get to know `R` and RStudio, we'll be working with the `R` package [`leaflet`](https://rstudio.github.io/leaflet/). `leaflet` is the `R` implementation of [`leaflet.js`](http://leafletjs.com), an open-source Java Script library for building interactive maps.

### A Simple Map

`leaflet` itself is straightforward to get up and running. If we wanted an interactive map with a marker placed on-top of Chapin Hall, we would use the following script entered into `R`:

```{r map-chapin}
leaflet() %>%
  addTiles() %>%
  addMarkers(lng = -87.59267827941107, lat = 41.785848099872744, popup="Chapin Hall")
```

The `leaflet()` function creates a map widget, and the `addTiles()` function adds a base map to it. By default, [OpenStreetMap](https://www.openstreetmap.org) is used for the base map. Finally, we use `addMarkers()` to specify the longitude and latitude of our marker, and we enter in a label that will appear as a pop-up when a user clicks on the marker. `lng`, `lat`, and `popup` are all called "arguments" - these are used to control how a function operates.

The `%>%` is called the "pipe operator", and it is used to chain together functions in what we will call "pipelines". This pipeline can be read like a list, with the word **then** substituted for each instance of `%>%`:

1. First we create a map widget, **then**
2. we add base map tiles, **then**
3. we add a marker at the given longitude and latitude.

### Changing the Base Map

To alter the base map, we can use `addProviderTiles()` in place of `addTiles()`. I like the CartoDB "Positron" base map. To use the Positron base map, we create a second pipeline:

```{r map-chapin-carto}
leaflet() %>%
  addProviderTiles() %>%
  addMarkers(lng = -87.59267827941107, lat = 41.785848099872744, popup="Chapin Hall")
```

Two things are important to note here. When we load the `leaflet` package, we have access to a data object called `providers`. You can use the following syntax to explore it:

```{r explore}
names(providers)
```

`providers` is a vector of items, each of which corresponds to a different base map. We can select one of those items, `CartoDB.Positron`, by separating `providers` from the item name with a dollar sign (`$`). This is a classic way in which elements of a data set are accessed in `R` syntax.

Now, you try to select a base map from `providers` and alter the code chunk from above to reproduce the map showing Chapin Hall:

```{r map-chapin-alt}
leaflet() %>%
  addProviderTiles() %>%
  addMarkers(lng = -87.59267827941107, lat = 41.785848099872744, popup="Chapin Hall")
```

### Adding Additional Points

The `data/woodlawn_homicide.csv` file (a `.csv` file is a type of spreadsheet) contains information on recent homicides in Chicago's Woodlawn neighborhood. We can read it into `R` using the `readr` package (part of the tidyverse):

```{r read-woodlawn-data}
woodlawn <- read_csv(here("data", "woodlawn_homicide.csv"))
```

We read the statement from right to left - the data found at `data/woodlawn_homicide.csv` is read correctly as `.csv` data, and the resulting imported data is stored in an object in our global environment named `woodlawn`. The `here()` function helps us write simple, operating system agnostic file paths that will always be relative to where the `.Rproj` file is stored. We'll talk more about this as the semester progresses.

We can explore the data a number of ways, including with the `View()` (output not shown) function and the `str()` function:

```{r explore-woodlawn}

```

If we wanted to use `View()`, it would be implemented like this:

```r
View(woodlawn)
```

When executed in the console, it will produce a spreadsheet-like view within RStudio.

The `.csv` data are *tabular* data - they contain longitude and latitude data, but they are not *projected*. This means we are missing the geometric data that locates these longitude and latitude data in space. leaflet can take these spatial references, however, and convert them to usable geometric data. We do so using a very similar process to what we did before:

```{r map-places}

```

The `data = woodlawn` argument in `leaflet()` directs `R` to the appropriate data set to map. We use the tilde (`~`) to indicate to leaflet that these are variables within `woodlawn`.

## Converting to `sf` Objects

In practice, we don't usually build maps this way. Instead, we convert our tabular data to an `sf` object, which contains a `geometry` column that is ready for mapping and spatial data wrangling. We'll load a larger data set of homicides to illustrate this process:

```{r convert-homicides}
## load data
homicide <- read_csv(here("data", "chicago_homicide.csv"))

## project data

```

The `longitude` and `latitude` columns contain our `x` and `y` values for coordinates, respectively. Longitude lines are the vertical lines that extend east and west from the prime meridian. Latitude lines are the horizontal lines that extend north and south from the equator. 

The `crs` value of `4269` refers to a *geographic coordinate system,* which is a model of the earth's surface that expresses locations in decimal degrees. These are another way of representing longitude and latitude. `4269` is known as "NAD83," which stands for North American Datum 1983. This is a model that is specific to North America. The other common geographic coordinate system we use is WGS84, which can be used anywhere in the world. Its `crs` value is `4326`.

Once we build `homicide_sf`, we can see that its properties have changed:

```{r class-sf}

```

We can also easily start to work with our data. For example, we can use `View()` (or click on our object in the global environment) to get a sense of its columns and rows. Note the `geometry` column - this contains the information `R` packages need to create maps with our data.

We can also use the `mapview` package to preview our data:

```{r preview-homicide}

```

This is particularly useful for making sure data have been projected correctly, and getting a sense of the spatial *extent* of our data.

Now, you repeat the process of loading and projecting data using the file `data/chicago_arson.csv`, which contains locations of recent arson fires in Chicago. Be sure to check your variable names, as they are different than in the `homicide` data.

```{r process-arsons}
## load data


## project data


## preview data

```

## Prepping Our Data
In order to map our data properly with `leaflet`, they must use the WGS 84 *geographic coordinate system*. As we noted above, the EPSG code for WGS 84 is `4326`. To make our transformations, we'll use `st_transform()` paired with our selected EPSG code:

```{r transform-homicides}

```
 
Now, you try this syntax out on the demographic data:

```{r transform-arson}

```

## Creating Interactive Maps
Finally, we'll build some simple interactive maps. Since our data are `sf` objects, we no longer need to specify the `latitude` and `longitude` columns. This makes our code a bit easier to read and write! We use the homicide data's incident date combined with a little html code to make our pop-up:

```{r map-homicides}

```

Now, you repeat this process for the arson data, swapping out out both the base map, the data object, and any variable names that may have changed:

```{r map-arson}

```
